Solving the Nonstationary Richards Equation\\ With Adaptive $hp$-FEM

### 2011-03-15 paper.tex
Fixing some stuff.

\begin{abstract}
%% Text of abstract
This paper examines the potential of the adaptive $hp$-FEM method for the 
numerical solution of time-dependent variably saturated 
Darcian flow problems described by the Richards equation. The method is 
illustrated on three model problems: a benchmark with known exact solution, 
groundwater seepage into a dry lysimeter box with time-dependent boundary 
conditions, and capillary barrier behavior under an intense infiltration. 
In the second part of the paper we present a brief overview of adaptive 
$hp$-FEM and introduce the open source adaptive $hp$-FEM library HERMES.
All computations presented in this paper are easily reproducible.
\end{abstract}

@@

%tady jsem konkretni cisla section zmenil na odkaz na sekci, vysazene to vypada stejne
The outline of the paper is as follows: To make the paper self-contained and to 
establish notation, in Section~\ref{sec:richards} we introduce the Richards equation and 
relevant constitutive relations. In Section~\ref{sec:hpfem} we
demonstrate the potential of adaptive $hp$-FEM using an elliptic benchmark problem with known 
exact solution. In Section~\ref{sec:compar}, an analogous comparison with low-order finite element 
methods is done for the time-dependent Richards equation, in the context of a 
benchmark problem with known exact solution. Section~\ref{sec:example} is dedicated to solving 
with adaptive $hp$-FEM a nonstationary groundwater seepage into a dry lysimeter box,
with time-dependent boundary conditions. Section~\ref{sec:example-barrier} deals with 
a nonstationary problem of capillary barrier behavior under an intense infiltration into 
an initially unsaturated medium. Weak forms for the Richards equation are presented in 
Section \ref{sec:8} and conclusions and outlook are drawn in Section \ref{sec:9}.

In Appendix A we describe in more detail some particular aspects on adaptive 
$hp$-FEM that are relevant for computations performed in this paper, and we 
introduce the open source project HERMES \cite{hermes}.

@@

%odkomentoval jsem, protoze se na tuto analytickou rovnici budu dal v textu odkazovat
When van Genuchten's formula (\ref{vangenuchten-eq}) is substituted in to this integral 
law, the following analytical formula is obtained

@@

The finite element method (FEM) is the most widely used numerical technique
to solve partial differential equations (PDE). 
By adaptive $hp$-FEM we mean a modern version of FEM that combines
adaptively finite elements of different spatial diameters $(h)$ and polynomial
degrees $(p)$ in order to maximize the convergence rate.
For theory and technical details of higher-order FEM
and automatic $hp$-adaptivity we refer, e.g., to \cite{sosedo} 
and many references therein. In the following let us present
a simple example that illustrates the huge computational potential of 
the method.

@@

Since the treatment of the initial condition (\ref{ic}) is nontrivial
in the finite element context, let us say a few words about it. 
At time $t = 0$ the exact solution (\ref{ic}) is constant and it 
does not satisfy the Dirichlet boundary condition on the top edge. 
Therefore we begin at a small positive time $t = 10^{-3}$.
It is accepted that discontinuous initial condition is extremely common
in practice and thus it will appear in the last example (\ref{sec:example-barrier}).

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[height=4.5cm]{mesh_init_h1.png}
\end{center}
\vspace{-6mm}
\caption{Initial mesh for $h$-FEM with linear elements (24843 DOF).}
\label{fig:init-meshes-3}
\end{figure}
Implicit Euler method was used for time integration.
In Figs. \ref{fig:init-meshes-1} - \ref{fig:init-meshes-3} 
different colors are used to distinguish between 
different polynomial degrees of finite elements. All elements in these computations 
are quadrilaterals. Notice in Fig. \ref{fig:init-meshes-1} that adaptive $hp$-FEM employs
anisotropic (i.e., directionally different) polynomial degrees in finite 
elements -- this is when an element in the mesh contains two different colors. 
The steady-state solution is shown in Fig. \ref{fig:steady-state}.

@@

It is well known  that
both the van Genuchten's model~(\ref{vangenuchten-eq}) and Mualem's model~(\ref{mualem-anal})
involves a large amount of computer operations for its evaluation (see e.g. Ross~\cite{ross} or Simunek et. al.~\cite{simunek}).
Methods to approximate the unsaturated hydraulic conductivity function were discussed by Miller et al.~\cite{miller}
and Hermite polynomials were recommended to use.

The function $C(h)$ was replaced by Hermit cubic splines (only its first derivative is required 
by the Newton's method) and $K(h)$ was replaced with Hermit quintic spline (its first and second derivative is required 
by the Newton method). The derivatives are obtained by derivating the spline functions.
Via this approximation we cured the problem of singularity of the second derivative 
of $K(h)$,

@@

The previous sections~\ref{sec:compar} -- \ref{sec:example-barrier} presents an application
 of the Richards' equation law on various theoretical and engineer problems. An h-based form~(\ref{rich-final})  numerically approximated by $h$ and $hp$ adaptive finite element method was considered. It is well known (see e.g. Celia, et. al.~\cite{celia}) that improper time discretization and integration techniques of this form of the Richards' equation exhibit in mass balance errors and erroneous estimates of an infiltration depth. 
The error is caused by an improper discrete approximation of the time derivative term $C(h)\frac{\partial h}{\partial t}$, originating from mass (water content) derivative term $\frac{\partial \theta(h)}{\partial t}$.
Higher order time integration methods as considered by Kees et al.~\cite{kees-time-integ} for two-phase flow can treat the mass balance problem. An adaptivity in time as a response to changing conditions is an efficient approach to maintain mass conservation.
Alternatively, in case of a low order time integration method,  Kuraz~et.al.~\cite{kuraz-jcam} recommends time step adaptivity criterion based on retention curve torsion.

The current implementation gives an estimate on error in  spatial approximation, nevertheless the time integration is maintained by 
low order Crank-Nicholson and implicit Euler scheme. The implementation of the Richards equation into the Hermes project is currently  under 
a development and the Crank-Nicholson scheme is not the final solution. More robust method, a so call SDIRKNG (Singly diagonally implicit Runge-Kutta-Nystr\'om  general), see~\cite{sdirk}, is under an evaluation. The time integration will be based on arbitrary Butcher's tables and then the time adaptivity will be enabled by embedded Runge-Kutta's methods by estimating truncation error.

Based on the results obtained from all conducted examples it might be apparent that the $hp$ adaptive method requires a very similar computational CPU demands compared to $h$ adaptivity with $p=2$. The advantage of $hp$ FEM methods appears together with an increasing demands for spatial accuracy due to its exponential convergence. As input data uncertainty is a typical issue for many hydrological problems $h$ adaptivity with $p=2$ can provide a sufficiently good performance for many engineering applications.

\section{Weak Formulation for the Newton's and Picard's Methods}\label{sec:9}

Let $\Omega$ be a bounded domain in $\mathbb{R}^2$,
with a Lipschitz boundary $\partial\Omega$. Let
$\Gamma_D$ and $\Gamma_N$ be smooth open disjoint subsets of
$\partial\Omega$ (not necessarily connected) such that
$\partial\Omega = \overline{\Gamma_D}\cup\overline{\Gamma_N}$,
$\textmd{meas}_{n-1}(\partial\Omega\setminus({\Gamma_D}\cup{\Gamma_N}
))=0$ and $\textmd{meas}_{n-1}(\Gamma_D)>0$ and
$\textmd{meas}_{n-1}(\Gamma_N)>0$. In what follows $\bfn$ denotes the outer unit
normal to $\partial\Omega$. For positive $T$ let
$Q_T=\Omega\times[0,T)$.

Let
\begin{displaymath}
{\mathcal{E}}(\overline{\Omega}):=\left\{\bfv\in{\mathcal
C}^\infty(\overline{\Omega})^n; \, {\textmd{supp}\, \bfv}  \cap
\Gamma_D = \emptyset \right\}.
\end{displaymath}
Let the linear space $\mathbb{V}$ be a closure of
${\mathcal{E}}(\overline{\Omega})$ in the norm of $W^{1,2}(\om)$.

The mathematical model of unsaturated flow with initial data and the
mixed boundary conditions is described by the following
initial--boundary value problem
\begin{align}
 C(h)\frac{\partial h}{\partial t}
 - \nabla \cdot \left( K(h) \nabla h \right)
  -  \frac{\partial K(h)}{\partial z}
  &= 0  &&\textmd{in}\; Q_T,
\label{dual_richards}
\\
h(0) &= h^0 &&\textmd{in}\;\Omega, \label{dual_richards_init}
\\
h &= h_D  &&\textmd{on}\;\Gamma_D \times
[0,T),\label{dirichlet_dual}
\\
\frac{\partial h}{\partial \bfn} &= g &&\textmd{on}\;\Gamma_N \times
[0,T). \label{neumann1_dual}
\end{align}

Here we consider all functions to be smooth enough and that the
compatibility condition $h^0=h_D$ on $\Gamma_D$ holds.

The weak formulation of the problem
\eqref{dual_richards}--\eqref{neumann1_dual} is as follows:

Let the given data satisfy
\begin{equation}
h^0 \in W^{1,2}(\Gamma_D),  \; h^0 = h_D(0) \; \textmd{on}\;\Gamma_D,
\end{equation}
\begin{equation}
\begin{split}
g \in \mathcal{C}^1([0,T];L^{\infty}(\Gamma_N)),  \; h_D \in
\mathcal{C}^1([0,T];  \\ L^{\infty}(\Gamma_D)\cap W^{1,2}(\Omega)).
\end{split}
\end{equation}
A function
\begin{equation*}
\begin{split}
h \in \mathcal{C}([0,T);W^{1,2}(\Omega))\cap
\mathcal{C}([0,T); \\ L^{\infty}(\Omega))\cap
\mathcal{C}^1([0,T);L^{2}(\Omega)),
\end{split}
\end{equation*}
such that
$$
h(0)=h_0 \textmd{ in } \Omega, \; h(t)=h_D(t) \textmd{ on } \Gamma_D
\times [0,T),
$$
is a variational solution to the system
\eqref{dual_richards}--\eqref{neumann1_dual} iff
\begin{equation}
\begin{split}
\int_{\Omega}  C(h)\frac{\partial h}{\partial t}\varphi \;{\rm
d}\bfx + \int_{\Omega} K(h)\nabla h \cdot \nabla \varphi \;{\rm
d}\bfx - \\ - \int_{\Omega}\frac{\partial K(h)}{\partial z}\varphi \;{\rm
d}\bfx = \int_{\Gamma_N} g \varphi \;{\rm d}\bfS
\end{split}
\label{discrete-rich}
\end{equation}

holds for all $\varphi \in \mathbb{V}$ and $t\in(0,T)$.

The time discretization is accomplished for the example in section~\ref{sec:example} through a Crank-Nicholson scheme, then in  a compact form the  problem~\eqref{discrete-rich} can be written as

\begin{equation}
\begin{split}
\mathbf{C(h}_n^{}) \frac{\mathbf{h}_{n}-\mathbf{h}_{n-1}}{\Delta t} +  \frac{1}{2}(\mathbf{K(h}_n^{})\mathbf{h}_n + \mathbf{K'(h}_n^{})\mathbf{h}_n  \\ \mathbf{K(h}_{n-1}^{})\mathbf{h}_{n-1} + \mathbf{K'(h}_{n-1}^{})\mathbf{h}_{n-1}) +  \mathbf{f} = 0 
\end{split}
\end{equation}

where subscript $n$ denotes the time step number, $\Delta t$ denotes the time step length and $\mathbf{f}$ denotes the boundary value data.

The examples in sections~\ref{sec:compar} and~\ref{sec:example-barrier} consider time integration method through an implicit Euler scheme, then in a compact form the  problem~\eqref{discrete-rich} can be written as

\begin{equation}
\begin{split}
\mathbf{C(h}_n^{}) \frac{\mathbf{h}_{n}-\mathbf{h}_{n-1}}{\Delta t} +  \mathbf{K(h}_n^{})\mathbf{h}_n + \mathbf{K'(h}_n^{})\mathbf{h}_n +  \mathbf{f} = 0 
\end{split}
\end{equation}

Because of the non-linearity of the Richards equation model both the Picard iteration method and the Newton iteration method was implemented to obtain the solution. Matrix $\mathbf{K'(h)}$ is obtained  from derivatives of $K(h)$ function in respect to $h$.
The Picard iteration procedure is formulated as

\begin{equation*}
\mathbf{C(h}_n^{i-1}) \frac{\mathbf{h}_{n}^i-\mathbf{h}_{n-1}}{\Delta t} +  \mathbf{K(h}_n^{i-1})\mathbf{h}_n + \mathbf{K'(h}_n^{i-1})\mathbf{h}_n^i +  \mathbf{f} = 0 
\end{equation*}

where superscript $i$ denotes the iteration level. The convergency criterion is defined as $||h^i - h^{i-1}|| < \varepsilon$, where $\varepsilon$ is some desired error.

The Newton procedure is formulated as

\begin{equation*}
\begin{split}
\frac{1}{\Delta t} \left( \frac{\partial }{\partial \mathbf{h}} \mathbf{C(h}_n^{i})(\mathbf{h}_n^{i}-\mathbf{h}_{n-1}) + \mathbf{C(h})_{n}^i \right)\Delta \mathbf{h}_n^i + \\ + \left(\frac{\partial }{\partial \mathbf{h}}
\mathbf{K(h)}_n^i\mathbf{h}^i_n + \mathbf{K}(\mathbf{h)}_n^i + \frac{\partial }{\partial \mathbf{h}}
\mathbf{K'(h)}_n^i\mathbf{h}^i_n + \mathbf{K'}(\mathbf{h)}_n^i \right) \Delta \mathbf{h}^i_n = \\ = - \left(\mathbf{C(h_n}^i)\frac{\mathbf{h}_n^i - \mathbf{h}_{n-1}}{\Delta t} + \mathbf{K(h)}_n^i\mathbf{h}^i_n + \mathbf{K'(h)}_n^i\mathbf{h}^i_n \right) ,
\end{split}
\end{equation*}

at the end of each iteration the primary variables are updated as follows

$$\mathbf{h}_n^{i+1} = \mathbf{h}_n^i + \Delta \mathbf{h}^i_n \quad .$$

>>>>

\begin{abstract}
%% Text of abstract
This paper examines the potential of the adaptive $hp$-FEM method for the 
numerical solution of time-dependent variably saturated 
Darcian flow problems described by the Richards equation. The method is 
illustrated on three model problems: a benchmark with known exact solution, 
groundwater seepage into a dry lysimeter box with time-dependent boundary 
conditions, and capillary barrier behavior under an intense infiltration. 
In the second part of the paper we present the weak formulation of the Richards equation 
for the Newton's and Picard's methods, give a brief overview of adaptive 
$hp$-FEM with emphasis on aspects that are usually not discussed in the 
literature, and we briefly introduce the open source adaptive $hp$-FEM library 
HERMES that was used to generate numerical results for this paper. All 
computations that we present are easily reproducible.
\end{abstract}

@@

%tady jsem konkretni cisla section zmenil na odkaz na sekci, vysazene to vypada stejne
The outline of the paper is as follows: To make the paper self-contained and to 
establish notation, in Section~\ref{sec:richards} we introduce the Richards equation and 
relevant constitutive relations. In Section~\ref{sec:hpfem} we
demonstrate the potential of adaptive $hp$-FEM using an elliptic benchmark problem with known 
exact solution. In Section~\ref{sec:compar}, an analogous comparison with low-order finite element 
methods is done for the time-dependent Richards equation, in the context of a 
benchmark problem with known exact solution. Section~\ref{sec:example} is dedicated to solving 
with adaptive $hp$-FEM a nonstationary groundwater seepage into a dry lysimeter box,
with time-dependent boundary conditions. Section~\ref{sec:example-barrier} deals with 
a nonstationary problem of capillary barrier behavior under an intense infiltration into 
an initially unsaturated medium. Conclusions and outlook are drawn in Section \ref{sec:8}.

In Appendix we present weak forms for the Richards equation and their solution 
via the Picard's and Newton's method, describe in more detail some particular aspects on adaptive 
$hp$-FEM that are relevant for computations performed in this paper. We also briefly 
introduce the open source adaptive $hp$-FEM library HERMES \cite{hermes} that was used to 
generate numerical results for this paper.

@@

%odkomentoval jsem, protoze se na tuto analytickou rovnici budu dal v textu odkazovat
When van Genuchten's formula (\ref{vangenuchten-eq}) is substituted into this integral 
law, the following analytical formula is obtained

@@

The finite element method (FEM) is the most widely used numerical technique
to solve partial differential equations (PDE). 
By adaptive $hp$-FEM we mean a modern version of FEM that combines
adaptively finite elements of different spatial diameters $(h)$ and polynomial
degrees $(p)$ in order to maximize the convergence rate.
For theory and technical details of higher-order FEM
and automatic $hp$-adaptivity we refer, e.g., to \cite{sosedo} 
and many references therein. In the following let us present
a simple example that illustrates the large computational potential of 
the method.

@@

Since the treatment of the initial condition (\ref{ic}) is nontrivial
in the finite element context, let us say a few words about it. 
At time $t = 0$ the exact solution (\ref{ic}) is constant and it 
does not satisfy the Dirichlet boundary condition on the top edge. 
Therefore we begin at a small positive time $t = 10^{-3}$.
We are aware of the fact that discontinuous initial conditions are common
in practice, and this is reflected in the last example (\ref{sec:example-barrier}).

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[height=4.5cm]{mesh_init_h1.png}
\end{center}
\vspace{-6mm}
\caption{Initial mesh for $h$-FEM with linear elements (24843 DOF).}
\label{fig:init-meshes-3}
\end{figure}
In this and all consequent examples we use the implicit second-order 
Crank-Nicolson method for time integration.
In Figs. \ref{fig:init-meshes-1} - \ref{fig:init-meshes-3} 
different colors are used to distinguish between 
different polynomial degrees of finite elements. All elements in these computations 
are quadrilaterals. Notice in Fig. \ref{fig:init-meshes-1} that adaptive $hp$-FEM employs
anisotropic (i.e., directionally different) polynomial degrees in finite 
elements -- this is when an element in the mesh contains two different colors. 
The steady-state solution is shown in Fig. \ref{fig:steady-state}.

@@

It is well known that
both the van Genuchten's model~(\ref{vangenuchten-eq}) and Mualem's model~(\ref{mualem-anal})
involve a large amount of computer operations (see, e.g., Ross~\cite{ross} or 
Simunek et. al.~\cite{simunek}). Various methods to approximate the unsaturated hydraulic 
conductivity function were discussed by Miller et al. in \cite{miller} who recommended
Hermite polynomials.

In our model, the function $C(h)$ was replaced by Hermite cubic splines (only its first derivative 
is required by the Newton's method) and $K(h)$ was replaced with Hermite quintic splinee (its first 
and second derivative are required by the Newton method). The derivatives are obtained by differentiating 
the spline functions, not by splining the derivatives.
Via this approximation we cured the problem of singularity of the second 
derivative of $K(h)$,

@@

We have examined the potential of adaptive $hp$-FEM
for the solution of the nonstationary Richards equation. 
The $h$-based form~(\ref{rich-final}) of this equation 
was solved by means of adaptive low-order 
and higher-order finite element methods. 
Based on the results of many numerical experiments, 
it is safe to say that adaptive $h$-FEM with quadratic elements $(p=2)$
is often competitive with adaptive $hp$-FEM when not very high spatial 
accuracy is required. If high accuracy in space matters, then 
the exponential convergence of the $hp$-FEM makes it more 
efficient than quadratic $h$-FEM. Notably, neither adaptive
method needed stabilization because singularities and steep
layers were resolved with sufficient accuracy.

It is well known that the $h$-based form~(\ref{rich-final}) 
of the Richards' equation is prone 
to errors in mass balance as well as to erroneous estimates of 
infiltration depth when inaccurate time integration is used (see, e.g., 
Celia et. al.~\cite{celia}). These errors are caused by improper discrete 
approximation of the time derivative term 
$C(h)\frac{\partial h}{\partial t}$ that originates in the mass (water content) derivative 
term $\frac{\partial \theta(h)}{\partial t}$. Kees et al.~\cite{kees-time-integ} have 
shown in the context of two-phase flow that higher-order time integration methods
can cure the mass balance problem.

The current study puts more emphasis on the spatial discretization methods 
and spatial adaptivity. Time integration is carried out using the 
second-order Crank-Nicolson method -- this is appropriate for quadratic $h$-FEM
but too low-order for adaptive $hp$-FEM. We are working on the implementation 
of advanced higher-order adaptive implicit Runge-Kutta methods and plan to 
publish a detailed study of this subject in near future.

\section{Weak Formulation of the Richard's Equation
for the Newton's and Picard's Methods}\label{sec:9}

Let $\Omega$ be a bounded domain in $\mathbb{R}^2$,
with a Lipschitz boundary $\partial\Omega$. Let
$\Gamma_D$ and $\Gamma_N$ be smooth open disjoint subsets of
$\partial\Omega$ (not necessarily connected) such that
$\partial\Omega = \overline{\Gamma_D}\cup\overline{\Gamma_N}$,
$\textmd{meas}_{n-1}(\partial\Omega\setminus({\Gamma_D}\cup{\Gamma_N}
))=0$ and $\textmd{meas}_{n-1}(\Gamma_D)>0$ and
$\textmd{meas}_{n-1}(\Gamma_N)>0$. In what follows $\bfn$ denotes the outer unit
normal to $\partial\Omega$. For positive $T$ let
$Q_T=\Omega\times[0,T)$.
Let
\begin{displaymath}
{\mathcal{E}}(\overline{\Omega}):=\left\{\bfv\in{\mathcal
C}^\infty(\overline{\Omega})^n; \, {\textmd{supp}\, \bfv}  \cap
\Gamma_D = \emptyset \right\}.
\end{displaymath}
Let the linear space $\mathbb{V}$ be a closure of
${\mathcal{E}}(\overline{\Omega})$ in the norm of $W^{1,2}(\om)$.
The mathematical model of unsaturated flow with initial data and the
mixed boundary conditions is described by the following
initial--boundary value problem
\begin{align}
 C(h)\frac{\partial h}{\partial t}
 - \nabla \cdot \left( K(h) \nabla h \right)
  -  \frac{\partial K(h)}{\partial z}
  &= 0  &&\textmd{in}\; Q_T,
\label{dual_richards}
\\
h(0) &= h^0 &&\textmd{in}\;\Omega, \label{dual_richards_init}
\\
h &= h_D  &&\textmd{on}\;\Gamma_D \times
[0,T),\label{dirichlet_dual}
\\
\frac{\partial h}{\partial \bfn} &= g &&\textmd{on}\;\Gamma_N \times
[0,T). \label{neumann1_dual}
\end{align}
We assume that all functions are smooth enough for the 
compatibility condition $h^0=h_D$ on $\Gamma_D$ to hold.
The weak formulation of the problem
\eqref{dual_richards}--\eqref{neumann1_dual} is as follows:
Let the given data satisfy
\begin{equation}
h^0 \in W^{1,2}(\Gamma_D),  \; h^0 = h_D(0) \; \textmd{on}\;\Gamma_D,
\end{equation}
\begin{equation}
\begin{split}
g \in \mathcal{C}^1([0,T];L^{\infty}(\Gamma_N)),  \; h_D \in
\mathcal{C}^1([0,T];  \\ L^{\infty}(\Gamma_D)\cap W^{1,2}(\Omega)).
\end{split}
\end{equation}
A function
\begin{equation*}
\begin{split}
h \in \mathcal{C}([0,T);W^{1,2}(\Omega))\cap
\mathcal{C}([0,T); \\ L^{\infty}(\Omega))\cap
\mathcal{C}^1([0,T);L^{2}(\Omega)),
\end{split}
\end{equation*}
such that
$$
h(0)=h_0 \textmd{ in } \Omega, \; h(t)=h_D(t) \textmd{ on } \Gamma_D
\times [0,T),
$$
is a variational solution to the system
\eqref{dual_richards}--\eqref{neumann1_dual} if
\begin{equation}
\begin{split}
\int_{\Omega}  C(h)\frac{\partial h}{\partial t}v \;{\rm
d}\bfx + \int_{\Omega} K(h)\nabla h \cdot \nabla v \;{\rm
d}\bfx \\ - \int_{\Omega}K'(h)\frac{\partial h}{\partial z}v \;{\rm
d}\bfx - \int_{\Gamma_N} g v \;{\rm d}\bfS = 0
\end{split}
\label{discrete-rich}
\end{equation}
holds for all $v \in \mathbb{V}$ and $t\in(0,T)$.

For simplicity, we will use the implicit Euler method 
to derive the discrete problems for the Picard's and 
Newton's methods. We replace the time-derivative
in \eqref{discrete-rich} with the backward time difference, 
the continuous unknown $h^{n+1}$ with a piecewise-polynomial function
$$
h^{n+1} = \sum_{j=1}^N y_j v_j
$$
and the (continuous) test function $v \in V$ with a finite 
element test function $v_i$ to obtain

\begin{equation}
\begin{split}
\int_{\Omega}  C(h^{n+1})\frac{h^{n+1} - h^n}{\Delta t}v_i \;{\rm
d}\bfx + \int_{\Omega} K(h^{n+1})\nabla h^{n+1} \cdot \nabla v_i \;{\rm
d}\bfx \\ - \int_{\Omega}K'(h^{n+1})\frac{\partial h^{n+1}}{\partial z}v_i \;{\rm
d}\bfx - \int_{\Gamma_N} g^{n+1} v_i \;{\rm d}\bfS = 0
\end{split}
\label{pavelfinal1}
\end{equation}
for all $i = 1, 2, \ldots, N$.
Here $N$ is the dimension of the finite element subspace of $V$,
the superscript $n$ enumerates time steps, $\Delta t$ is the length of the time step, 
and $\mathbf{f}$ stands for the contribution of the (generally time-dependent) 
Neumann boundary data.

Let us begin with the Picard's method. This method performs a simple linearization of 
nonlinear product terms by moving all of them but one to the previous iteration level.
For equation (\ref{pavelfinal1}) this means

\begin{equation}
\begin{split}
\int_{\Omega}  C(h_m^{n+1})\frac{h_{m+1}^{n+1} - h^n}{\Delta t}v_i \;{\rm
d}\bfx + \int_{\Omega} K(h_m^{n+1})\nabla h_{m+1}^{n+1} \cdot \nabla v_i \;{\rm
d}\bfx \\ - \int_{\Omega}K'(h_m^{n+1})\frac{\partial h_{m+1}^{n+1}}{\partial z}v_i \;{\rm
d}\bfx - \int_{\Gamma_N} g^{n+1} v_i \;{\rm d}\bfS = 0
\end{split}
\label{pavelfinal2}
\end{equation}
for all $i = 1, 2, \ldots, N$.
Here the subscript $m$ denotes the iteration level. As stopping criterion one can either 
use a residual obtained by inserting $h_{m+1}^{n+1}$ into (\ref{pavelfinal1}), or look at the 
distance of two consecutive Picard's approximations $\Vert h_{m+1}^{n+1} - h_{m}^{n+1} \Vert$.
Equation (\ref{pavelfinal2}) constitutes a system of linear algebraic equations to be solved
in each iteration.

For the Newton's method it is useful to view the piecewise-polynomial
function $h^{n+1}$ as a function of the coefficient vector $Y = (y_1, y_2, \ldots, y_N)^T$,
$$
h^{n+1} = \sum_{j=1}^N y_j v_j = h(Y).
$$
Then the left-hand side in (\ref{pavelfinal1}) can be viewed as a nonlinear 
function of the coefficient vector $Y$ as well. This function is the $i$th component 
of the vector-valued nonlinear {\em residual} and we denote it by $F_i(Y)$.
Next we need to differentiate the function $F_i(Y)$ with respect to $y_j$ to obtain 
the Jacobian matrix entry $J_{ij}$,

$$
J_{ij}(Y) = \frac{\partial F_i(Y)}{\partial y_j} 
= \int_{\Omega}  C'(h(Y))v_j\frac{h(Y) - h^n}{\Delta t}v_i \;{\rm d}\bfx 
$$
$$
+ \int_{\Omega}  C(h(Y))\frac{v_j v_i}{\Delta t} \;{\rm d}\bfx 
+ \int_{\Omega} K'(h(Y))v_j\nabla h(Y) \cdot \nabla v_i \;{\rm d}\bfx 
$$
$$
+ \int_{\Omega} K(h(Y))\nabla v_j \cdot \nabla v_i \;{\rm d}\bfx 
- \int_{\Omega}K''(h(Y))v_j\frac{\partial h(Y)}{\partial z}v_i \;{\rm d}\bfx 
$$
$$
- \int_{\Omega}K'(h(Y))\frac{\partial v_j}{\partial z}v_i \;{\rm d}\bfx. 
$$
The Newton's method itself is then performed in the standard way,
$$
J(Y_m) \delta Y_{m+1} = -F(Y_m), \ \ \ Y_{m+1} = Y_{m} + \delta Y_{m+1}.
$$
